The Calculus of Linear Constructions

### 2021-12-02 writing/draft5/draft5.tex
writing
draft5

The presence of both linear and dependent types enables CLC to write specifications that faithfully encode the usage of resources. The previous example of monetary transaction can be refined using an indexed linear type family $Money : \mathbb{N} \rightarrow L$ as follows.
\begin{equation*}
  (x \ltype Money\ 5) \rightarrow (Camels + Marlboro)
\end{equation*}
This new specification for the transaction demands a payment of 5 units of money, and the customer be relieved of their ownership after the transaction finishes, effectively preventing the contradiction of having your cake and eating it too.

@@

\noindent \textbf{\textit{Contributions}}:
Our contributions can be summarized as follows.
\begin{itemize}
  \item Fist, we describe the Calculus of Linear Constructions, an extension to the Calculus of Constructions with linear types. The integration of linear types and dependent types allows CLC to directly and precisely reason about resource.
  \item Second, we study the meta-theory of CLC directly, showing that it satisfies the standard properties of confluence, validity and subject reduction.
  \item Next, we prove the linearity theorem, showing the tangible impact linear types have on the structure of terms. We also show that promotion and dereliction of linear logic are derivable as theorems for canonical types.
  \item Additionally, we observe that CLC is backwards compatible with the predicative CC$\omega$, allowing us to construct a reduction preserving model of CLC in CC$\omega$, showing that CLC is consistent. We prove that all valid CC$\omega$ terms are valid CLC terms after adding minimal amounts of annotation.
  \item Furthermore, we extend CLC with rules to define inductive linear types in the style of CIC, forming the Calculus of Inductive Linear Constructions (CILC). This is the first formalization of a generalized mechanism for defining inductive linear types that we are aware of.
  \item All major results have been formalized and proven correct in the Coq Proof Assistant with help from the Autosubst \cite{autosubst} library. To the best of our knowledge, our development is the first machine checked formalization of a linear dependent type theory.
  \item Finally, we give an implementation of CILC in OCaml. Algorithmic type checking employed by the typechecker streamlines the process of writing CILC programs.
\end{itemize}

@@

\subsection{Internalization of Effect Laws}
The combination of linear types and dependent types within the state axioms enable the internalization of the so called "effect laws" for state, where these laws are now provable theorems within the CLC framework.

@@

\begin{figure}[h]
  \caption{Law 1}
  \begin{minted}{coq}
  let [ _, c ] := new _ 1 in
  let [ xeq, c ] := get _ _ _ c in
  let [ yeq, c ] := get _ _ _ c in
  let ( x, pf1 ) := xeq in
  let ( y, pf2 ) := yeq in
  let pf1 := Eq_sym _ _ _ pf1 in
  let pf : Eq _ x y := 
    Eq_trans _ _ _ _ pf1 pf2 
  in free _ _ _ c.
  \end{minted}
  \label{law1}
  \Description{}
\end{figure}

@@

\begin{figure}[h]
  \caption{Law 2}
  \begin{minted}{coq}
  let [ _, c ] := new _ 1 in
  let c := set _ _ _ _ c 2 in
  let [ zeq, c ] := get _ _ _ c in
  let ( z, pf1 ) := zeq in
  let pf : Eq _ 2 z := pf1 in
  free _ _ _ c.
  \end{minted}
  \label{law2}
  \Description{}
\end{figure}

\subsection{Protocol Enforcement}
The inclusion of dependent types and linear types in CILC makes it extremely expressive when encoding communication protocols. By indexing each channel with a protocol, CLC can enforce communication on a channel to strictly adhere to its specified protocol.

>>>>

The presence of both linear and dependent types enables CLC to write specifications that faithfully encode the usage of resources. The previous example of monetary transaction can be refined using an indexed linear type family $Money : \mathbb{N} \rightarrow L$ as follows.
\begin{equation*}
  (x \ltype Money\ 5) \rightarrow (Camels \oplus Marlboro)
\end{equation*}
This new specification for the transaction demands a payment of 5 units of money, and the customer be relieved of their ownership after the transaction finishes, effectively preventing the contradiction of having your cake and eating it too.

@@

\noindent \textbf{\textit{Contributions}}:
Our contributions can be summarized as follows.
\begin{itemize}
  \item Fist, we describe the Calculus of Linear Constructions, an extension to the Calculus of Constructions with linear types. The integration of linear types and dependent types allows CLC to directly and precisely reason about resource.
  \item Second, we study the meta-theory of CLC directly, showing that it satisfies the standard properties of confluence, validity and subject reduction.
  \item Next, we prove the linearity theorem, showing the tangible impact linear types have on the structure of terms. We also show that promotion and dereliction of linear logic are derivable as theorems for canonical types.
  \item Additionally, we observe that CLC is backwards compatible with the predicative CC$\omega$, allowing us to construct a reduction preserving model of CLC in CC$\omega$, showing that CLC is consistent. We prove that all valid CC$\omega$ terms are valid CLC terms after adding minimal amounts of annotation.
  \item Furthermore, we extend CLC with rules to define inductive linear types in the style of CIC, forming the Calculus of Inductive Linear Constructions (CILC). This is the first presentation of a generalized mechanism for defining inductive linear types that we are aware of.
  \item All major results have been formalized and proven correct in the Coq Proof Assistant with help from the Autosubst \cite{autosubst} library. To the best of our knowledge, our development is the first machine checked formalization of a linear dependent type theory.
  \item Finally, we give an implementation of CILC in OCaml. Algorithmic type checking employed by the typechecker streamlines the process of writing CILC programs.
\end{itemize}

@@

The combination of linear types and dependent types within the state axioms enable the internalization of the so called "effect laws" for state, where these laws are now provable theorems within the CILC framework.

@@

\begin{figure}[h]
  \caption{Law 1}
  \begin{minted}{coq}
  let [ _, c ] := new _ 1 in
  let [ xeq, c ] := get _ _ _ c in
  let [ yeq, c ] := get _ _ _ c in
  let ( x, pf1 ) := xeq in
  let ( y, pf2 ) := yeq in
  let pf1 := eq_sym _ _ _ pf1 in
  let pf : eq _ x y := 
    eq_trans _ _ _ _ pf1 pf2 
  in free _ _ _ c.
  \end{minted}
  \label{law1}
  \Description{}
\end{figure}

@@

\begin{figure}[h]
  \caption{Law 2}
  \begin{minted}{coq}
  let [ _, c ] := new _ 1 in
  let c := set _ _ _ _ c 2 in
  let [ zeq, c ] := get _ _ _ c in
  let ( z, pf1 ) := zeq in
  let pf : eq _ 2 z := pf1 in
  free _ _ _ c.
  \end{minted}
  \label{law2}
  \Description{}
\end{figure}

\subsection{Protocol Enforcement}
The inclusion of dependent types and linear types in CILC makes it extremely expressive when encoding communication protocols. By indexing each channel with a protocol, CILC can enforce communication on a channel to strictly adhere to its specified protocol.
